---
title: "The logic behind if_date() and restrict_date()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The logic behind if_date() and restrict_date()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Case definitions often have a temporal condition, e.g.,

> two or more physician visits **at least 30 days apart within two years**.

`if_date()` (for a vector of dates) and `restrict_date()` (for data frames and database tables) decide whether a client's dates can satisfy such a condition. This vignette explains how they work under the hood: why the `within` part is easy, why the `apart` part needs a search algorithm, and how that algorithm gets translated into SQL — a language with no loops — using metaprogramming and 'dbplyr'.

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(healthdb)
```

## Why not brute force?

The question to answer is: among a client's dates, **is there any set of `n` dates** whose adjacent gaps are all at least `apart` days *and* whose overall span is at most `within` days?

The direct approach is to try every possible set of `n` dates:

```{r}
brute_force <- function(x, n, apart = 0, within = Inf) {
  any(utils::combn(
    unique(x), n,
    function(s) all(diff(sort(s)) >= apart) & (max(s) - min(s) <= within)
  ))
}
```

This is correct, and it is exactly what the package's tests compare against. But the number of sets explodes: a client with 20 dates has `choose(20, 3)` = `r choose(20, 3)` sets of three, and `choose(20, 5)` = `r format(choose(20, 5), big.mark = ",")` sets of five. Across millions of clients this is painful in R — and impossible on a database, because SQL has no way to enumerate combinations at all. Both conditions need a smarter route.

## The easy half: `within`

If the only condition is "any `n` dates within `within` days", sorting solves it. After sorting, **the `n` dates with the smallest span are always `n` consecutive dates** — skipping over a date in between can never shrink a span. So instead of all combinations, we only check each run of `n` consecutive sorted dates: is the gap between a date and the (`n` − 1)-th date after it at most `within`?

```{r}
x <- as.Date("2020-01-01") + c(0, 200, 380, 390)

# any 3 of these within one year? check consecutive triples:
# (0, 200, 380) spans 380; (200, 380, 390) spans 190 -- yes
if_date(x, n = 3, within = 365)
```

This is a rolling window over sorted dates — one pass, no combinations. On a database, the same idea is one window function: compare each date with `lead(date, n - 1)` (or `lag()`, depending on `flag_at`) within the client.

## The hard half: `apart`

Now the condition "any `n` dates pairwise at least `apart` days apart". Sorting alone does not help here: the qualifying dates need not be consecutive — in fact they are usually spread out, and there are combinatorially many ways to pick a spread-out set. We need a search, but one that avoids enumerating sets.

### Anchor at the extremes

The key observation: **if any qualifying set exists, then a qualifying set that uses the client's earliest and latest dates also exists.**

Why? Take any qualifying set and swap its earliest member for the client's overall earliest date. That swap moves the first pick further away from (or keeps the same distance to) the second pick, so the first gap can only grow — the set still qualifies. The same argument applies at the other end with the latest date.

This means we never have to wonder which dates to start and end with: anchor at the two extremes, and the problem shrinks to finding the **middle** `n − 2` dates. And the middle dates can't be just anywhere — each must be at least `apart` days after the overall minimum and at least `apart` days before the overall maximum. In other words, they must fall inside the window

> [ min + apart,  max − apart ].

Inside that window, the same logic repeats: anchor at the window's own earliest and latest dates, shrink the window by `apart` on both sides, and look for `n − 4` dates in the new, smaller window. Every pass peels two dates off `n`, so the search finishes in `n %/% 2` passes — regardless of how many dates the client has:

-   For **odd** `n`, the last pass needs to find just **one** date in the innermost window.
-   For **even** `n`, the last two picks are the innermost window's own earliest and latest dates, so the final check is whether **their** gap is at least `apart`.
-   At every pass, if the window holds fewer than the required `n − 2i` dates, the answer is no — the search stops early.

### A worked example

Take a client whose dates are days 0, 5, 9, 16, 23, and 30 (relative to their first visit), and ask for `n = 4` dates at least 7 days apart:

```{r}
x <- as.Date("2020-01-01") + c(0, 5, 9, 16, 23, 30)
```

-   **Pass 1**: anchor at day 0 and day 30. The middle two dates must lie in [0 + 7, 30 − 7] = [7, 23]. Days 9, 16, and 23 do — at least 2, so continue.
-   **Pass 2**: `n` is now down to 0, so apply the even-`n` ending: the earliest and latest dates inside the window are days 9 and 23, and 23 − 9 = 14 ≥ 7. **Qualifies** — the witness set is {0, 9, 23, 30} with gaps 9, 14, 7.

```{r}
if_date(x, n = 4, apart = 7)
```

Now shift two dates so the middle collapses — days 0, 5, 9, 12, 30:

-   **Pass 1**: window [7, 23] contains days 9 and 12 — still 2 dates, continue.
-   **Pass 2**: their gap is 12 − 9 = 3 < 7. **Fails.** No other choice could have done better: anything outside [7, 23] is too close to an anchor.

```{r}
y <- as.Date("2020-01-01") + c(0, 5, 9, 12, 30)
if_date(y, n = 4, apart = 7)
```

### Both conditions at once

When `apart` and `within` are both supplied, the two tricks combine. Any qualifying set spans at most `within` days, so it must live entirely inside the `within`-day window that starts at its own earliest date — and that earliest date is one of the client's dates. So:

1.  for each date, gather all the client's dates falling within `within` days of it (an *overlap join* — `data.table::foverlaps()` locally, a non-equi `left_join()` on the database);
2.  run the `apart` search from the previous section inside each of these windows;
3.  the client qualifies if any window does.

This is also where the `flag_at`/`align` argument gets its meaning: it decides whether the window anchored at a date extends forward ("left", the date is the start of the period) or backward ("right", the date is the end), which in turn decides which record carries the flag.

## Teaching SQL to loop

The `apart` search is naturally a loop: *for pass i, compute window i from the dates that survived window i − 1*. R can write that loop in a few lines, and `data.table` runs it for the local method. But `restrict_date()` also has to run on database tables, and SQL has no `for` loop.

The way out: **the loop does not have to run in SQL — it only has to run while *writing* the SQL.** Each pass of the search is expressible as ordinary 'dplyr' verbs on grouped data: window `i` is a `mutate()` that uses `min()`, `max()`, and `sum()` over the client group. So the package loops in R *at query-building time*, manufacturing one pair of columns per pass with 'rlang' metaprogramming:

```{r, eval=FALSE}
## simplified from healthdb's internals
for (i in 1:(n %/% 2)) {
  win_i <- rlang::sym(paste0("in_win_", i)) # column name for this pass
  win_prev <- rlang::sym(paste0("in_win_", i - 1))
  sum_i <- rlang::sym(paste0("sum_win_", i))

  data <- data %>%
    mutate(
      # is the date inside this pass's window (computed from the
      # dates that were inside the previous pass's window)?
      !!win_i := case_when(
        between(
          date,
          min(date[!!win_prev == 1L]) + apart,
          max(date[!!win_prev == 1L]) - apart
        ) ~ 1L,
        .default = 0L
      ),
      # are there enough dates left in the window?
      !!sum_i := case_when(sum(!!win_i) >= n - i * 2L ~ 1L, .default = 0L)
    )
}
```

`rlang::sym()` turns a generated name like `"in_win_2"` into a column reference, and `!!` injects it into the `mutate()`. The loop runs `n %/% 2` times in R, leaving behind a chain of `mutate()` calls — a fully *unrolled* version of the loop. 'dbplyr' then translates each `mutate()` into a layer of SQL: the aggregate functions become window functions (`MIN(...) OVER (PARTITION BY clnt_id)` and friends), and each pass becomes a nested subquery.

You can see the generated SQL for yourself:

```{r}
db <- dbplyr::memdb_frame(
  clnt_id = 1,
  uid = 1:6,
  dates = as.Date("2020-01-01") + c(0, 5, 9, 16, 23, 30)
)

restrict_date(db, clnt_id, dates, n = 4, apart = 7, uid = uid) %>%
  show_query()
```

Reading from the innermost subquery outward, you can follow the passes of the search: `in_win_1` (the first window), `sum_win_1` (enough dates left?), then the final gap check — exactly the worked example above, written by a loop SQL never sees. Asking for a larger `n` simply generates more layers; no R code changes.

This pattern — looping in R to generate set-based SQL — is worth knowing beyond this package. Any per-group algorithm with a bounded number of passes, where each pass is expressible as `mutate()`/`summarise()` over the group, can be unrolled the same way and pushed to the database instead of downloading the data.

## Trust, but verify

Because the brute force version is unquestionably correct, it makes a good referee. The package's tests compare both implementations against it; here is the flavor:

```{r}
set.seed(44)

referee <- replicate(200, {
  x <- sample(seq(as.Date("2020-01-01"), as.Date("2021-12-31"), by = 1), 8)
  identical(
    if_date(x, n = 3, apart = 30, within = 365),
    brute_force(x, n = 3, apart = 30, within = 365)
  )
})

all(referee)
```

The same checks run for the database method against SQLite, PostgreSQL, and SQL Server, so the SQL that the metaprogramming writes is held to the same standard as the R that is easy to read.

## Where this sits among known techniques

For readers who want to relate the algorithm to established ideas (and for the record of what we could and could not find elsewhere):

**The selection problem itself is classical — when solved sequentially.** "Pick `n` points with pairwise gaps at least `g`" has a textbook greedy solution: sort the dates, take the earliest, then repeatedly take the next date at least `g` after the last pick, and check whether `n` picks were reached. It is linear after sorting, and its correctness rests on an *exchange argument* — the same style of proof used in the anchoring step above (competitive programmers may recognize the feasibility check inside the "aggressive cows" / maximize-the-minimum-gap problem). The catch: this greedy is inherently sequential — each pick depends on the previous one — which makes it a poor fit for SQL. Expressing it would require recursion (recursive common table expressions), which is not portable across database dialects and not supported by 'dbplyr'.

**The shrinking-window formulation removes the sequential dependency.** Each pass needs only `min()`, `max()`, and a count over the client's group — all ordinary window aggregates — and the number of passes is fixed at `n %/% 2` before seeing any data. That is what makes the algorithm expressible in plain, non-recursive SQL. The nearest well-known SQL patterns solve different problems: [gaps-and-islands](https://www.practicewindowfunctions.com/learn/gap_and_island) classifies *consecutive* rows into runs, and `LEAD()`/`LAG()` threshold checks (as used for the `within` condition here) also only compare *consecutive* rows — but a qualifying `apart` set is generally not consecutive. Academic work on [threshold queries](https://arxiv.org/abs/2106.15703) studies a related family of "are there at least n rows such that ..." questions, but not this gap-constrained subset selection. We have not found a published set-based equivalent of the shrinking-window search; pointers are welcome on the [issue tracker](https://github.com/KevinHzq/healthdb/issues).

**The condition itself is everywhere in epidemiology, but implemented sequentially.** Validation studies routinely define cases by patterns like "two or more claims at least 30 days apart" — for example, in [case-finding algorithms for HIV](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3557280/) and [chronic dialysis](https://link.springer.com/article/10.1186/1471-2288-11-25) — and the published implementations are typically SAS or Stata programs that scan records row by row. Running the same logic as one set-based query inside the database appears to be uncommon.

**Generating SQL from a host language is an old idea; this use of it is not well documented.** Query builders and 'dbplyr' itself are code generators, and the 'rlang' injection mechanics are covered in the ['dbplyr' translation vignette](https://dbplyr.tidyverse.org/articles/sql-translation.html) and community posts such as [this one on passing functions to 'dbplyr'](https://nathaneastwood.github.io/2021/02/18/using-functions-as-an-input-to-functions-with-dbplyr/). What we have not seen documented elsewhere is the pattern used here: running a loop at query-building time to unroll a bounded-pass, per-group algorithm into nested window-function subqueries, as a general way to push iterative logic to the database.
