---
title: "Intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

------------------------------------------------------------------------

## What it does {#sec-what-it-does}

-   Designed for identifying disease cases from admin data for epidemiological studies. The implementation focused on code readability and re-usability.

-   Interactive functions (e.g., `identify_rows()`, `exclude()`, `fetch_vars()`) based on filter and joins from dplyr with tweaks that fix SQL translation or add features that are not natively support by SQL. They also work for local data.frame, and some use `data.table` package to speed up processing time for large data. These functions are not as flexible as dplyr::filter, but they can be applied to analyses outside health research.

-   Call-building functions (e.g., `build_def()`, `execute_def()`) that facilitate batch execution and re-use of disease case definitions. In essence, `build_def` creates codes of definitions (which is chain of the interactive functions, e.g., `define_case()`) that are not ran immediately but can be evaluated with different input data through `execute_def`.

-   Miscellaneous functions such as computing age, collapsing records within a time range into one episode, and more (on-going effort) with built-in checks signalling if things could go wrong.

### Motivation

Administrative health data data are often stored on database with strict security measures which may disable permission to write temporary tables. Writing queries without being able to cache intermediate results is challenging, especially when the data is too large to be downloaded from database into R (i.e., local memory) without some filtering process.

The `dbplyr` package not only translates `dplyr` syntax to SQL but also make nesting sub-queries easy with the pipe operator. With dbplyr, complex query script can be built from re-using sub-queries instead of writing temporary tables at a cost of performance (sub-queries have to be ran again and again if it was called multiple times). Nevertheless, building complex queries with dbplyr is a useful approach to push computation onto the SQL server that is often much more efficient and resourceful than the user's (virtual) machine.

However, dbplyr SQL translations failed at some places for the intended use case and needs fixing.

------------------------------------------------------------------------

## Installation {#sec-installation}

Simply run:

```{r, eval = FALSE}
devtools::install_github("kevinHzq/odcfun", build_vignettes = TRUE)
```

We will also need the following packagesfor this demo.

```{r setup, message=FALSE}
library(dplyr)
library(dbplyr)
library(lubridate)
library(glue)
library(purrr)
library(odcfun)
```

## Intended use case {#sec-intended-use-case}

Consider the case definition of substance use disorder (SUD) from [British Columbia Centre for Disease Control's Chronic Disease Dashboard](http://www.bccdc.ca/resource-gallery/Documents/Chronic-Disease-Dashboard/substance-use-disorder.pdf),

> One or more hospitalization with a substance use disorder diagnostic code, OR Two or more physician visits with a substance use disorder diagnostic code within one year.

We are going to implement this definition. First, let's make a demo data sets for the two sources:

1.  Physician claims with multiple columns of [ICD-9](https://www2.gov.bc.ca/gov/content/health/practitioner-professional-resources/msp/physicians/diagnostic-code-descriptions-icd-9) diagnostic codes

    ```{r}
    #make_test_dat() makes either a toy data.frame or database table in memory with known number of rows that satisfy the query we will show later 
    claim_db <- make_test_dat(vals_kept = c("303", "304", "305", "291", "292",glue("30{30:59}"), glue("29{10:29}"), noise_val = c("999", "111")), type = "database")

    #this is a database table
    #note that in-memory SQLite database stores dates as numbers
    claim_db %>% head()
    ```

2.  Hospitalization with [ICD-10](https://en.wikipedia.org/wiki/ICD-10) codes

    ```{r}
    hosp_df <- make_test_dat(vals_kept = c(glue("F{10:19}"), glue("F{100:199}"), noise_val = "999"), type = "data.frame")

    #this is a local data.frame/tibble
    hosp_df %>% head()
    ```

## Interactive functions

Let's focus on the physician claims. Extracting clients with at least two records within a year is not difficult, and involves only a few steps. The codes could look like the following using dplyr, however, it does not work because: 1. SQL does not support multiple patterns in one LIKE operation, 2. dbply currently have issue with translating n_distinct.

```{r, eval=FALSE}
## not run
claim_db %>% 
  #identify the target codes
  filter(if_any(starts_with("diagx_"), ~ str_like(., c("291%", "292%", "303%", "304%", "305%")))) %>%
  #each clnt has at least 2 records on different dates
  group_by(clnt_id) %>%
  #the n_distinct step is mainly for reducing computation in the next step
  filter(n_distinct(dates) >= 2) %>% 
  #any two dates within one year?
  filter((max(dates) - min(dates)) <= 365)
## end
```

Here's how you could use `odcfun` to achieve these steps:

1.  Identify rows contains the target codes

    ```{r}
    result1 <- claim_db %>%
      identify_rows(vars = starts_with("diagx_"),
                    match = "start",
                    vals = c(291:292, 303:305))
    ```

2.  Bonus: remove clients with exclusion codes

    This step is not in the substance use disorder definition, but other disease definitions often require exclusion of some ICDs that contradicts the ones of interest. Let's say we want to remove clients with code "111" here.

    We first identify "111" from the source, then exclude clients in the output from the previous step's result. \`exclude\` take either a data set (via the excl argument) or expression (condition argument) as input. For the former, it performs an anti join matching on the by argument (see dplyr::join_by). For the latter, it is the opposite of filter, i.e., filter(!(some_expression)).

    ```{r}
    result2 <- result1 %>% 
      exclude(excl = identify_rows(claim_db, starts_with("diagx_"), "in", "111"), 
              by = "clnt_id")
    ```

3.  Restrict the number of records per client

    ```{r}
    result3 <- result2 %>% restrict_n(clnt_id = clnt_id, 
                                      n_per_clnt = 2, 
                                      count_by = dates,
                                      # here we use filter mode to remove records that failed the restriction
                                      mode = "filter"
                                      )
    ```

4.  Restrict the temporal pattern of diagnoses

    \`restrict_dates\` also supports more complicated patterns like having n diagnoses at least i days apart within j years, but the "apart" feature requires relatively expensive computation and implemented for local data.frames only. Note that when SQL interpret order of dates, the result could be not deterministic if there were duplicate dates within client. Therefore, a unique row id colume (uid) has to be supplied to get consistent result.

    ```{r}
    result4 <- result3 %>% restrict_dates(clnt_id = clnt_id,
                                          date_var = dates,
                                          n = 2, 
                                          within = 365,
                                          uid = uid,
                                          # here we use flag mode to flag records that met the restriction instead of removing those
                                          mode = "flag")
    ```

5.  Fetch variables from other tables by matching common keys

    Up to this point, the result is only a query and have not been downloaded. Hopefully, it has been shrunken to a manageable size for collection.

    ```{r}
    # Class of result4
    class(result4)

    # execute query and download the result
    result_df <- result4 %>% collect()

    # Number of rows in source
    nrow(claim_db %>% collect())

    # Number of rows in the current result
    nrow(result_df)

    ```

    Our data now only contains diagnoses that are probably not enough for further analyses. Let's say we want to gather client demographics such as age and sex from other sources. This certainly can be done with multiple `dplyr::left_join` calls. Here we provide the `fetch_var` function to make the codes more concise.

    ```{r}
    # make two look up tables
    age_tab <- data.frame(clnt_id = 1:50, 
                          age = sample(1:90, 50),
                          sex = sample(c("F", "M"), 50, replace = TRUE))
    address_tab <- data.frame(clnt_id = rep(1:50, 5), year = rep(2016:2020, each = 50),
      area_code = sample(0:200, 50, replace = TRUE))

    # get year from dates for matching 

    result_df <- result_df %>% mutate(year = lubridate::year(as.Date(dates, origin = "1970-01-01")))

    #note that keys must be present in all tables
    result_df %>% 
      fetch_vars(keys = c(clnt_id, year),
                 linkage = list(
                   # |clnt_id means matching on clnt_id only
                   age_tab ~ c(age, sex)|clnt_id,
                   address_tab ~ area_code
                 )) %>% 
      head()
    ```

## Call-building functions

To complete the definition, we need to repeat the process shown above with hospitalization data. Some studies may use more than a handful of data sources to define their sample. We packed steps 1-4 in one function `define_case`, and provide tools to perform batch execution with different data and parameters to meet those needs.

```{r}
# build the full definition of SUD
sud_def <- build_def(
  #name of definition
  def_lab = "SUD",
  #place holder names for sources
  src_labs = c("claim", "hosp"),
  def_fn = define_case, # you could alter it and supply your own function
  #below are argumets of define_case
  fn_args = list(
    # if length = 1, the single element will be use for every source 
    vars = list(starts_with("diagx_")),
    match = "start", # match ICD starts with vals
    vals = list(c(291:292, 303:305), glue("F{10:19}")),
    clnt_id = clnt_id,
    n_per_clnt = c(2, 1),
    date_var = dates,
    within = c(365, NULL),
    uid = uid,
    mode = "flag" 
  )
)

sud_def
```

Let's look inside the fn_call list column. Two calls of `define_case` have been made with different parameters. The data arguments are left empty on purpose for re-usability. For example, you may want to repeat the analysis with data from different regions or study periods.

```{r}
sud_def$fn_call
```

Executing the definition is simple. If verbose option is not turned off by `options(odcfun.verbose = FALSE)`, the output message will explain what has been done. You could append multiple `build_def` outputs together and execute them all at once. Definition and source labels will be added to the result to identify outputs from different calls.

```{r}
# execute the definition
result_list <- sud_def %>% 
  execute_def(with_data = list(
    claim = claim_db,
    hosp = hosp_df
  ))

# view the results
purrr::walk(result_list, ~ head(.) %>% print)
```
